\act

\scene{ --- A design bureau at 9:30 AM}

Employees arriving.  Drinking coffee.  Sitting down. Busily looking at screens.  Some small sounds come from occasional design actions 

Eliza, a slightly nerdy looking young woman enters from the back of the stage, holding a portfolio.  She is wearing a tacky, oversized sweatshirt with fred flintstone (or similar) on it, a large bow in her hair, white sneakers and glasses that went out of style 10 years ago.

She stands for several minutes watching the members of the company work unsure about approaching them.


\dialog{Raster} Who the fuck is the charity store refugee lurking in the corner?

\dialog{Kern} Oh my! Are they shooting an early 90s period film about color blind sports fans?

\dialog{Raster} You are good!

\dialog{Raster} So what the hell does she want?

Hilde motions to Liza to come over.  They have a short whispered exchange.  Hilde looks skeptical.

\dialog{Hilde} Um.  Apparently, she wants to apply to be our new intern.

Raster shreiks and nearly falls out of his chair.

\dialog{Raster} WHAT!?  Oh fuck no!

\dialog{Kern} Well I guess I should never have given up my childhood dreams of dancing at the Bolshoi.

\dialog{Raster} Tell her to sit down.

\dialog{Liza} Hi!  I rlly want to be ur intern.  U are sriously pro designers! :)

\dialog{Raster} Could you maybe start by telling us your name.

\dialog{Liza} Oh right srry.  Nervous!  I'm Liza.  Liza Doolittle.

\dialog{Raster} Raster.  I assume, if you know anything at all, you have heard of me.

\dialog{Liza} mm.  yeah.  somebody said ur super good.

\dialog{Kern} Nice to meet you Liza.  My name is Kern.

\dialog{Raster} So what makes you think we would want you at all?  Where did you study?  RCA?

\dialog{Liza} no...

\dialog{Raster} RISD?  Universitaet der Kuenste?  Willem de Kooning?

\dialog{Liza} no erm.  i havent been to college.

\dialog{Kern} Oh I see.  So you are self-taught?

\dialog{Liza} yeah!  I work in a copy shop.  I use free software like gimp and mypaint.

took a night course in ArtySuite TS-9. And I got a book for christmas.  Learn to make killer PhotoStore version 12.666 images in just 24 hours. 

\dialog{Raster} The depravity.  Shocking.

\dialog{Kern} Hmm.. never heard of those.  So you don't use any software from the Stucco Corporation?

\dialog{Liza} I can't afford it.  

\dialog{Raster} Well then you are not a real designer.  We only use the absolute latest versions of Arty Suite CS 12.

\dialog{Kern} In the old days I did some rather good work with one of those horrible photo-setting machines and a drafting table.  So how about showing us your portfolio?

\dialog{Raster} No way.  Don't waste my fucking time.

\dialog{Kern} Henry, be nice.  Liza can you just give us a quick overview?

\dialog{Liza} ok

SOLO: slide show of her very funky / tacky / prolo / dirt-style portfolio 
Raster talks trash the whole time.  Moaning, sighing.  Oh my eyes! etc.
Kern is more generous but obviously not impressed.

\dialog{Liza}{so, what do you think?}

\dialog{Raster} I have no use for you.

\dialog{Liza} ill work for free!  ill clean the bathrooms.  just teach me to be a real designer.

\dialog{Kern} Henry, weren't you saying a few days ago that you could teach anyone to pass for a serious designer? madison avenuei <- photoshop jockey

\dialog{Raster} Yes, well...

\dialog{Kern} Well, lets say we have a little bet.  If she can give a presentation at the Art Dictators Club meeting in three months and be well recieved, I will pay you 50 bitcoins.

\dialog{Raster} Oh you tempt me.  Ok!  You're on!  Liza, you are a complete mess of a human being but we will tidy out up a bit.  But once it's over you are out.  

\dialog{Liza} Really!  OMG!  thks so much.  I will work rlly hard.  promise.

\dialog{Raster} Yes.  Hilde, take Liza in the back and burn her clothes.

\dialog{Hilde} Ok.  But what's she going to wear then?

\dialog{Raster} Wrap her in a blanket and have the Benneton store deliver some new outfits.  Strictly black and grey.  Oh and get her an iPhone too.


The assistant is sent off with eliza and told to burn her clothes and order a new outfit from Benneton, strictly black and grey.
