% Demo file for using ScriptTex 1.03%version

% This isn't a particularly exciting script, but it is demonstrative of what
% you can do with ScriptTex.

\input screenplay.tex

% optional \twoside
% optional \jlbrooks

\TV % or \movie (\TV is default)
\proof % or \manuscript or \shoot  (\proof is default)

\series{}  % Optional
\title{Demonstration Title} % used on title page and in \proof footlines
\author{Adrian C. McCarthy} % used on title page and in \proof footlines
\basedon{Adapted from the novel \em{For Example}}
\draft{First draft} % used only if \proof is in force
\date{25-OCT-1991}  % defaults to \today
\showcopyright{\number\year}
\address{1234 Main Street \\
         Anytown, MO 54321 \\
         (323) 555-1234}

\maketitle % creates a title page

%%%%
%%%% Main script begins here.
%%%% For the first part of this script, I won't take advantage of any synonyms
%%%% or shortcuts.
%%%%

\teaser % begins a teaser act

\open{fade in}

\scene{ext.\ forest~--- morning}

\stage The sun is just barely climbing over the tops of the pine trees.
We hear the \sound{roar} of a chainsaw.

\dialog{Lumberjack}
\paren{off screen}
Timber!

\stage One of the tall pines begins to give, and then, with a loud
\sound{crash}, it falls.

\trans{cut to}

\scene{ext.\ road nearby}

\stage A stationwagon approaches, parks, and \introduce{Bob Bush} climbs out. 
He's wearing boots, a brown suit, and a hardhat.  A \extra{Foreman} talks with
him momentarily.

\introduce{Sam Savvy} approaches, and the foreman walks away in disgust.

\dialog{Sam} \paren{to Bob} What's his problem?

\dialog{Bob} Don't pay any attention to him.

%%%% A weak example of \leadin and \pickup:

\dialog{Sam} \leadin{You're gonna have to cut back}

\dialog{Bob} \pickup cut back on our quotas?

\dialog{Sam} We're under a lot of pressure from the lobbyists.

\dialog{Bob} You EPA guys have no backbone.

%%%% Let's make this long enough to demonstrate the page break handling of
%%%% ScriptTex.

\dialog{Sam} No backbone?  No backbone! \paren{fuming} Well you've got some
nerve.  Every year you cut down thousands of trees that take hundreds of years
to grow.  Sure, for each you kill, you plant a new one, but at the current
rate, there will be a shortage of trees while we wait for these saplings to
grow.  On top of that, many of these great trees are wasted on low-grade pulp
for cardboard and paper products.  If you would supply all of the necessary
pulp from the low quality trees, that would greatly reduce the problem.  Now
you say that we EPA-types have no backbone!  We've got the coconuts to stand up
and say, ``Wait a minute; we're destroying the planet.  Let's stop and find a
better way.''  You don't have the guts to try something new, something that may
even make you more money in the long run than cowering away from your social
responsibilities.

\close{fade out}

%%%%
%%%% The next chunk will show off some of the synonyms and shortcuts.
%%%%

\act % ends the teaser and starts the first act.

\fadein

\scene{Bob's breakfast nook~--- Sunday morning}

The sports page is open and on the card table; the rest of the paper is
stacked off to the side, unread.  Bob's in some grungy old sweats, playing
solitaire as he reads.

The phone \sound{rings}.

Bob answers.

\def\Bob{\dialog{Bob}}

\Bob Hello? \beat Yeah.  OK, I'm on my way.

He hangs up, and drops the rest of the cards onto the table.  He slips off the
sweats, and he's wearing his brown suit underneath.  He checks his reflection
in the side of the toaster, then heads out the door.

\dissolve

\scene{Bob's car}

Bob is driving way too fast.  To keep from mumbling to himself, he
turns on the radio.

\dialog{Announcer} \vo There's a tie-up on the Logjam ramp near the interstate.

\Bob Damn!

\fadeout
\theend

%%%% Always end with \bye.
\bye % cleans up and exits
